New cardinality estimation algorithms for HyperLogLog sketches

### 2016-08-07 paper/paper.tex
various changes:
* refactored c++ source code, new class for joint statistic of 2 hyperloglogs
* improved intersection calculation
* some paper improvements

% SYMBOLS
\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexK}{k}
\symDefine{\symIndexL}{l}
\symDefine{\symIndexJ}{j}
\symDefine{\symCount}{c}
\symDefine{\symCountVariate}{C}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symKappa}{\kappa}
\symDefine{\symX}{x}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}

@@

@@

Under the assumption of a uniform hash function, the probability that the register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ are equal to $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ is given by the corresponding probability mass function
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots,\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexI=1}^\symNumReg \symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that the smallest of $\symCardinality$ uniformly distributed random integers in range $[0, 2^\symRegRange)$ has a binary representation with its first 1-bit at position $\symRegVal$,
\begin{equation}
\symRegProbability(\symRegVal\vert\symCardinality) 
:=
\begin{cases}
1 & \symCardinality=0 \wedge \symRegVal = 0\\
0 & \symCardinality=0 \wedge 1\leq\symRegVal\leq\symRegRange+1\\
0 & \symCardinality\geq1 \wedge \symRegVal = 0\\
\left(1-\frac{1}{2^\symRegVal}\right)^\symCardinality - \left(1-\frac{1}{2^{\symRegVal-1}}\right)^\symCardinality & \symCardinality \geq 1 \wedge 1\leq\symRegVal\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1
\end{cases}.
\end{equation}

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean (HM) was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator which is given by
\begin{equation}
\label{equ:raw_estimator}
\symCardinalityRawEstimate
=
\alpha_\symNumReg
\operatorname{HM}
\!\left(
\symNumReg 2^{\symRegValVariate_1},\ldots,\symNumReg2^{\symRegValVariate_\symNumReg}
\right)
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexI=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexI}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal 2^{-\symRegVal}}.
\end{equation}
Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\!\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{\!\!-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the estimation error.

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality. The chart is based on 10\,000 randomly generated HyperLogLog sketches. More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}. Obviously, the raw estimator is biased for small and large cardinalities where it fails to return accurate estimates. In order to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_estimate}
\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:raw_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of 10\,000 randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
\label{fig:small_range_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{original_estimate}
\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of 10\,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:original_estimate}
\end{figure}

@@

\subsection{Estimation error}
\label{sec:corrected_raw_estimation_error}
In order to verify the new estimation algorithm, we generated 10\,000 HyperLogLog sketches and inserted up to 50 billion unique elements. Assuming a uniform hash function, element hash values can be mocked by random numbers. For the following results we used the Mersenne Twister random number generator with 19\,937 bit state size from the C++ standard library.

@@

We are not the first applying the maximum likelihood method to HyperLogLog sketches \cite{Clifford2012}. However, there a different element insertion algorithm was considered, which is very inefficient in practice due to its $\symBigO(\symNumReg)$ runtime complexity. Each register has its own hash function which is applied to the inserted element. The result is used to update the corresponding register. The advantage is that the register values are statistically independent in this case, which allows factorization of the joint probability distribution of all register values.

In contrast, \cref{alg:insert} uses what is also called stochastic averaging. Instead of iterating through all registers, a single hash value is used, to select a single register and to update its value. Since the register values are statistically dependent in this case, the maximum likelihood method is much more complicated to apply. However, with the help of the Poisson approximation we are finally able to derive a new robust and efficient cardinality estimation algorithm. Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the raw estimator \eqref{equ:raw_estimator} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.

@@

Using the probability mass function of the Poisson model \eqref{equ:poisson_pmf} the log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCountVariate_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCountVariate_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ is the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCountVariate_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.45\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be infinite.

@@

\subsection{Maximum likelihood estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of the computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCountVariate_\symRegVal$ is non-zero. If we denote $\symRegValVariate_\text{min}:=\min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$ and $\symRegValVariate_\text{max}:=\max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
\item The sum $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ in \eqref{equ:funcprime} can be precalculated and reused for all function evaluations.
\item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$. This series corresponds to a geometric series with ratio 2. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we know $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper(4\symX) = \frac{\symX+\symHelper(2\symX)\left(1-\symHelper(2\symX)\right)}{\symX+\left(1-\symHelper(2\symX)\right)}.
\end{equation}
%or
%\begin{equation}
%\label{equ:helper_recursion2}
%\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
%\end{equation}
This recursive formula is stable in a sense that the relative error of $\symHelper(4\symX)$ is smaller than that of $\symHelper(2\symX)$ as shown in \cref{app:helper_stable}.

>>>>

% SYMBOLS
\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexK}{k}
\symDefine{\symIndexL}{l}
\symDefine{\symIndexJ}{j}
\symDefine{\symCount}{c}
\symDefine{\symCountVariate}{C}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symKappa}{\kappa}
\symDefine{\symX}{x}
\symDefine{\symPhi}{\varphi}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symHelperW}{w}
\symDefine{\symHelperV}{v}
\symDefine{\symHelperU}{u}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}

@@

The desribed element insertion algorithm makes use of what is known as stochastic averaging \cite{Flajolet1985}. Instead of updating each of all $\symNumReg$ registers using $\symNumReg$ independent hash values, which would be an $\symBigO(\symNumReg)$ operation, only one register is selected and updated, which requires only a single hash function and reduces the complexity to $\symBigO(1)$.

@@

Under the assumption of a uniform hash function, the probability that the register values $\boldsymbol{\symRegValVariate} = (\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ are equal to $\boldsymbol{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ is given by the corresponding probability mass function
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\boldsymbol{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots,\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexI=1}^\symNumReg \symRegProbability(\symRegVal_\symIndexI\vert\symCardinality_\symIndexI)
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability(\symRegVal\vert\symCardinality)$ is the probability that the value of a register is equal to $\symRegVal$, after it was selected $\symCardinality$ times by the insertion algorithm
\begin{equation}
\symRegProbability(\symRegVal\vert\symCardinality) 
:=
\begin{cases}
1 & \symCardinality=0 \wedge \symRegVal = 0\\
0 & \symCardinality=0 \wedge 1\leq\symRegVal\leq\symRegRange+1\\
0 & \symCardinality\geq1 \wedge \symRegVal = 0\\
\left(1-\frac{1}{2^\symRegVal}\right)^\symCardinality - \left(1-\frac{1}{2^{\symRegVal-1}}\right)^\symCardinality & \symCardinality \geq 1 \wedge 1\leq\symRegVal\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symCardinality & \symCardinality\geq 1 \wedge \symRegVal = \symRegRange +1
\end{cases}.
\end{equation}

@@

In the history of the HyperLogLog algorithm different averaging techniques have been proposed. First, there was the LogLog algorithm using the geometric mean and the SuperLogLog algorithm that enhanced the estimate by truncating the largest register values before applying the geometric mean \cite{Durand2003}. Finally, the harmonic mean was found to give even better estimates as it is inherently less sensitive to outliers. The result is the so-called raw estimator which is given by
\begin{equation}
\label{equ:raw_estimator}
\symCardinalityRawEstimate
=
\alpha_\symNumReg
\frac{\symNumReg}
{\frac{1}{\symNumReg 2^{\symRegValVariate_1}}+\ldots+\frac{1}{\symNumReg2^{\symRegValVariate_\symNumReg}}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symIndexI=1}^{\symNumReg}2^{-\symRegValVariate_\symIndexI}}
= 
\frac{\symAlpha_\symNumReg \symNumReg^2}{\sum_{\symRegVal=0}^{\symRegRange+1}\symCountVariate_\symRegVal 2^{-\symRegVal}}.
\end{equation}
Here $\alpha_\symNumReg$ is a bias correction factor which was derived for a given number of registers $\symNumReg$ to be \cite{Flajolet2007}
\begin{equation}
\symAlpha_\symNumReg := \left(
\symNumReg
\int_0^\infty
\left(
\log_2\!\left(
\frac{2+u}{1+u}
\right)
\right)^\symNumReg
du
\right)^{\!\!-1}.
\end{equation}
Numerical approximations of $\symAlpha_\symNumReg$ for various values of $\symNumReg$ have been listed in \cite{Flajolet2007}. These approximations are used in many HyperLogLog implementations. However, since the published constants have been rounded to 4 significant digits, these approximations even introduce some bias for very high precisions $\symPrecision$. For HyperLogLog sketches that are used in practice with 256 or more registers ($\symPrecision\geq 8$), it is completely sufficient to use 
\begin{equation}
\label{equ:alpha_inf}
\symAlpha_\infty := \lim_{\symNumReg\rightarrow\infty} \symAlpha_\symNumReg = \frac{1}{2\log 2} \approx 0.7213475,
\end{equation}
as approximation for $\symAlpha_\symNumReg$ in \eqref{equ:raw_estimator}, because the additional bias is negligible compared to the estimation error.

\cref{fig:raw_estimate} shows the distribution of the relative error for the raw estimator as function of the cardinality. The chart is based on 10,000 randomly generated HyperLogLog sketches. More details of the experimental setup will be explained later in \cref{sec:corrected_raw_estimation_error}. Obviously, the raw estimator is biased for small and large cardinalities where it fails to return accurate estimates. In order to cover the entire range of cardinalities, corrections for small and large cardinalities have been proposed.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_estimate}
\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of 10,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:raw_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of 10,000 randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
\label{fig:small_range_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{original_estimate}
\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of 10,000 randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:original_estimate}
\end{figure}

@@

\subsection{Estimation error}
\label{sec:corrected_raw_estimation_error}
In order to verify the new estimation algorithm, we generated 10,000 HyperLogLog sketches and inserted up to 50 billion unique elements. Assuming a uniform hash function, element hash values can be mocked by random numbers. For the following results we used the Mersenne Twister random number generator with 19,937 bit state size from the C++ standard library.

@@

We remark that we are not the first applying the maximum likelihood method to HyperLogLog sketches \cite{Clifford2012}, where the case without stochastic averaging (compare \cref{sec:data_element_insertion}) was considered. The register values are statistically independent by nature in this case which 
allows factorization of the joint probability mass function and allows a straightforward application of the maximum likelihood method.

The more relevant case with stochastic averaging, which is considered in this paper, is more complicated. However, with the help of the Poisson approximation, we are able to still apply the maximum likelihood method and to derive a new robust and efficient cardinality estimation algorithm. Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the raw estimator \eqref{equ:raw_estimator} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.

@@

Using the probability mass function of the Poisson model \eqref{equ:poisson_pmf} the log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCountVariate_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCountVariate_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ is the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCountVariate_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.6\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be infinite.

@@

\subsection{Maximum likelihood estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of the computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCountVariate_\symRegVal$ is non-zero. If we denote $\symRegValVariate_\text{min}:=\min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$ and $\symRegValVariate_\text{max}:=\max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegValVariate_\text{min}, \symRegValVariate_\text{max}]$.
\item The sum $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}$ in \eqref{equ:funcprime} can be precalculated and reused for all function evaluations.
\item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\left\lbrace\frac{\symX}{2^{\symRegValVariate'_\text{max}}},\frac{\symX}{2^{\symRegValVariate'_\text{max}-1}},\ldots,\frac{\symX}{2^{\symRegValVariate_\text{min}}}\right\rbrace$ where $\symRegValVariate'_\text{max} := \min(\symRegValVariate_\text{max}, \symRegRange)$. This series corresponds to a geometric series with ratio 2. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we know $\symHelper\!\left(\frac{\symX}{2^{\symRegValVariate'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper(4\symX) = \frac{\symX+\symHelper(2\symX)\left(1-\symHelper(2\symX)\right)}{\symX+\left(1-\symHelper(2\symX)\right)}.
\end{equation}
%or
%\begin{equation}
%\label{equ:helper_recursion2}
%\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\!\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
%\end{equation}
This recursive formula is stable in a sense that the relative error of $\symHelper(4\symX)$ is smaller than that of $\symHelper(2\symX)$ as shown in \cref{app:helper_stable}.
